---
title: "R_day3_Exercise"
author: "Kinam"
date: '2022-10-12'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Exercise 1

1. Use readRDS() to load to an object called
/t1-data/project/obds/shared/resources/2_r/my_day2matrix.rds
```{r}
matrix <- readRDS("/t1-data/project/obds/shared/resources/2_r/my_day2matrix.rds")
str(matrix)
dim(matrix)
```

2. Add row sums and means to this dataset
```{r}
matrix_col <- cbind(matrix, rowsum = rowSums(matrix), rowmean = rowMeans(matrix))
head(matrix_col)
```
3. Add col sums and means to this dataset
```{r}
matrix_row <- rbind(matrix_col, colsum = c(colSums(matrix), NA, NA), colmean = c(colMeans(matrix), NA, NA))
tail(matrix_row)
```


4. Load the ToothGrowth datset using data(ToothGrowth)
```{r}
data(ToothGrowth)
class(ToothGrowth)
dim(ToothGrowth)
```

5. How do find more information on this dataset?
```{r}
str(ToothGrowth)
```

6. What columns of data do we have?
```{r}
colnames(ToothGrowth)
```

7. What is the mean tooth length?
```{r}
mean(ToothGrowth$len)
```

8. What is the maximum and minimum of the tooth length?
```{r}
max(ToothGrowth$len)
min(ToothGrowth$len)
```

9. Can you calculate rowSums and rowMeans on this data? 
Ans. No, because supp column is a character (code below does throws an error)
```{r}
# rowSums(ToothGrowth)
# rowMeans(ToothGrowth)
```
## Exercise 2 - Sorting

1. Load the AirQuality dataset.
```{r}
data(airquality)
dim(airquality)
str(airquality)
```
2. Look up more information about this dataset.
```{r}
help(airquality)
```

3. Examine the dataset – how many rows and columns are there?
```{r}
dim(airquality)
```

4. Display the column headers.
```{r}
colnames(airquality)
```

5. Sort the dataframe by the first column (Ozone) in ascending order.
```{r}
airquality_sort <- airquality[order(airquality$Ozone), ]
head(airquality_sort)
```
6. Sort the dataframe by the Month and then the Temp column in descending order and save to a new variable.
```{r}
airquality_sort_new <- airquality[order(airquality$Month, airquality$Temp, decreasing = TRUE), ]
tail(airquality_sort_new)
```

7. Save this sorted data to a file on the server.
```{r}
write.table(airquality_sort_new, file = "airquality_sort.csv", sep = ",", quote = FALSE, row.names = F)
```

## Exercise 3 - Merging

1. Create the dataframes called buildings and data to merge:
buildings <- data.frame(location=c(1, 2, 3), name=c("b1", "b2", "b3"))
data <- data.frame(survey=c(1,1,1,2,2,2), location=c(1,2,3,2,3,1),
efficiency=c(51,64,70,71,80,58))
```{r}
buildings <- data.frame(location=c(1, 2, 3), name=c("b1", "b2", "b3"))
data <- data.frame(survey=c(1,1,1,2,2,2), location=c(1,2,3,2,3,1), efficiency=c(51,64,70,71,80,58))
head(buildings)
head(data)
```
2. What is the common key variable in these dataframes? 
Ans. Location
```{r}
colnames(data)
colnames(buildings)
```

3. Use the merge() function to merge the two dataframes by the common key variable into a new dataframe called buildingStats.
```{r}
buildingStats <- merge(buildings, data, by = "location", all = TRUE)
head(buildingStats)
```

## Exercise 4 - Aggregate

1. Aggregate (or group) the airquality data by Month and return means on each of the numeric variables. Also, remove “NA” values.
```{r}
airquality_month <- aggregate(airquality, by = list(Months = airquality$Month), FUN = mean, na.rm = T)
head(airquality_month)
```

2. Aggregate the Solar.R column by Month, returning means of Solar.R The header of column 1 should be Month. Remove “not available” values.
```{r}
solar_agg <- aggregate(airquality$Solar.R, by = list(Months = airquality$Month), FUN = mean, na.rm = T)
head(solar_agg)
```

3. Apply the standard deviation function to the data aggregation you have just done.
```{r}
solar_agg_sd <- aggregate(x = airquality$Solar.R, by = list(Months = airquality$Month), FUN = sd, na.rm = T)
head(solar_agg_sd)
```

*Use this command if you want it to keep the column name, otherwise it will name the column as x*
```{r}
aggregate(Solar.R~Month, airquality, FUN = mean, na.rm = T)
```
## Exercise 5 - Functions

1. Write a function to calculate the hypotenuse of a triangle given the length of the 2 sides.
Remember the hypotenuse is the square root of the sum of the squares - √(a² + b²)
Run the function you have created with different values.
```{r}
cal_hypotenuse <- function(side1, side2){
  hypo <- sqrt(side1^2 + side2^2)
  return(hypo)
}
```

*Function usage - Hypotenuse*
```{r}
cal_hypotenuse(7,8)
```

## Exercise 6 - apply

1. Create a numeric vector of length 10. Write an lapply and sapply statement to square each element. Compare the two outputs.
```{r}
square <- function(num) {
  sq <- num^2
  return(sq)
}
```

```{r}
vector <- 1:10
lapply(vector, square)
sapply(vector, square)
```

2. Generate a list of length 4 containing both numeric and logical (T/F) vectors. Write an lapply or sapply statement to calculate the sum of the elements in each vector.
```{r}
vector1 <- 1:10
vector2 <- c(T, F, T, F, T, T, F, T, F, F)
vector3 <- 10:200
vector4 <- c(T, F, F, F, T, T)
my_list <- list(vector1, vector2, vector3, vector4)
head(my_list)
```

```{r}
lapply(my_list, sum)
sapply(my_list, sum)
```

3. Write an sapply statement to repeat each element of each vector in your list three times e.g. 1, 4, 3 would become 1, 1, 1, 4, 4, 4, 3, 3, 3. Assign the output to a new list.
```{r}
vec1 <- c(1, 2, 3)
vec2 <- c(4, 6)
vec3 <- c(5, 10, 15)
list2 <- list(vec1, vec2, vec3)
head(list2)
```
```{r}
repeat_element <- function(vec){
  x <- rep(vec, each = 3)
  return(x)
}
```

```{r}
lapply(list2, repeat_element)
```

```{r}
new_list <- sapply(list2, repeat_element)
new_list
class(new_list)
```

## Exercise7 - Loops

1. Write a for loop that iterates over the numbers 1 to 7 and prints the cube of each number using print().
```{r}
for (i in 1:7) {
  print(i^3)
}
```

2. Write a for loop that iterates over the column names of the inbuilt iris dataset and print each together with the number of characters in the column name in parenthesis.
Example output: Sepal.Length (12).
Use the functions print(), paste0() and nchar(). Look up what these functions do using the help feature.
```{r}
data("iris")
dim(iris)
str(iris)
```
```{r}
for (i in colnames(iris)) {
  print(paste0(c(colnames(iris[i])), " ", "(" , nchar(colnames(iris[i])), ")"))
}
```

*Alternate way*
```{r}
for (i in colnames(iris)) {
  print(paste0(i, " ", "(" , nchar(i), ")"))
}
```

3. Write an ifelse loop to print the colours in colours_vector with four characters. Use nchar() colours_vector <- c("red", "orange", "purple", "yellow", "pink", "blue")
```{r}
colours_vector <- c("red", "orange", "purple", "yellow", "pink", "blue")
for (i in colours_vector) {
ifelse(nchar(i) == 4, print(i), print("color name is not 4 characters"))
}
```
```{r}
ifelse(nchar(colours_vector) == 4, colours_vector, "color name is not 4 characters")
```

```{r}
max("dog", "rabbit", "zoo")   #max - gives alphabetical in char vector
```


