---
title: "Template ggplot2"
author: "Kevin Rue-Albrecht"
date: "13/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise 1

1. Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

2. What does `ggplot(diamonds)` do?

```{r}
ggplot(diamonds)

```
```{r}
diamonds
```


> Answer:

## Add the aesthetics.

3. What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7)
  
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour = cut))
  
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour = cut)) +
  geom_smooth()


```

- Assign that last plot to an object called `obds_diamonds`.

```{r}
obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour = cut)) +
  geom_smooth()


```


# Exercise

Predict the difference between these two plots
*things defined in ggplot are known to whole layers, however, things defined in sepcific layers are known to only those layers*  - you can overwrite the things defined in ggplot for a specific layer by redefining it in that layer
### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:

# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).
```{r}
head(ChickWeight)
```

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point()
 
```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight)) +
 geom_point(aes(colour = Diet))

```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point(size = 0.7) +
  geom_smooth(method = lm)

```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point(size = 0.7) +
  geom_smooth(method = lm) +
  facet_wrap(~Diet)

```

- Assign that last plot to an object called `obds_chickweight`.

```{r}
obds_chickweight <- ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point(size = 0.7) +
  geom_smooth(method = lm) +
  facet_wrap(~Diet)

```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
msleep

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x = order)) +
  geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x = order)) +
  geom_bar()+
  theme(axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1))

```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x = order)) +
  geom_bar() +
  theme(
    axis.text.x = element_text(angle = 90, size = rel(0.9), hjust = 1, vjust = 0.3), 
    axis.title = element_text(size = rel(1.5))
    ) +
  labs(x = "Order", y = "Count")

```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus)) +
  theme(
    axis.text.x = element_text(angle = 90, size = rel(0.9), hjust = 1, vjust = 0.3), 
    axis.title = element_text(size = rel(1.5))
    ) +
  labs(x = "Order", y = "Count")

```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus)) +
  theme(
    axis.text.x = element_text(angle = 90, size = rel(0.9), hjust = 1, vjust = 0.3), 
    axis.title = element_text(size = rel(1)),
    legend.key.size = unit(2, "mm"),
    legend.text = element_text(size = rel(0.5))
    ) +
  labs(x = "Order", y = "Count") +
  guides(fill = guide_legend(ncol = 3))


```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus)) +
  theme(
    axis.text.x = element_text(angle = 90, size = rel(0.9), hjust = 1, vjust = 0.3), 
    axis.title = element_text(size = rel(1)),
    legend.key.size = unit(2, "mm"),
    legend.text = element_text(size = rel(0.5))
    ) +
  labs(x = "Order", y = "Count") +
  guides(fill = guide_legend(ncol = 3))

```

- Add a contour of thin black lines to the bars.

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus), colour = "black", size = 0.1) +
  theme(
    axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1), 
    axis.title = element_text(size = rel(1)),
    legend.key.size = unit(2, "mm"),
    legend.text = element_text(size = rel(0.5)),
    legend.justification = c("right", "top"),
    ) +
  labs(x = "Order", y = "Count", fill = "Genus") + #fill - for legend title
  guides(fill = guide_legend(ncol = 3))

```

- Assign that last plot to an object called `obds_msleep`.

```{r}

obds_msleep <- ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus), colour = "black", size = 0.1) +
  theme(
    axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1), 
    axis.title = element_text(size = rel(1)),
    legend.key.size = unit(2, "mm"),
    legend.text = element_text(size = rel(0.5)),
    legend.justification = c("right", "top"),
    ) +
  labs(x = "Order", y = "Count", fill = "Genus") + #fill - for legend title
  guides(fill = guide_legend(ncol = 3)) #this command is to fit the legend in 3 columns

```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=5, fig.width=7}
library(cowplot)
first_row <- cowplot::plot_grid(obds_diamonds, obds_chickweight, ncol = 2, nrow = 1)
second_row <- cowplot::plot_grid(first_row, obds_msleep, nrow = 2, align = "hv")
second_row  

```

```{r}
library(cowplot)
cowplot::plot_grid(obds_diamonds, obds_chickweight, obds_msleep, ncol = 2, nrow = 2, fig.height = 12, fig.width = 12)

```

Using `patchwork`.

```{r, fig.height=5, fig.width=7_}
library(patchwork)
(obds_chickweight+obds_diamonds)/obds_msleep

```

Using `gridExtra`.

```{r, fig.height=4, fig.width=12}
library(gridExtra)
grid.arrange(
  grobs = list(obds_diamonds, obds_chickweight, obds_msleep),
  widths = c(2, 0.25, 2, 5),
  layout_matrix = rbind(c(1, NA, 3),
                        c(4))
) #this puts the gap between the top two plots - the NA in the data matrix which in this case is "0.25 wide".


```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave("R_day4_ggplots.pdf", second_row, height = 8, width = 10)
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!
```{r}
mpg
```


```{r}
ggplot(mpg, aes(x = cty, y = hwy, color = cyl)) +
  geom_point() +
  facet_wrap(~manufacturer)
  



```

