---
title: "Template for R S3 object oriented programming"
author: "Kevin Rue-Albrecht"
date: "13/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Exercises

1. Character

```{r}
character_object <- "A"
character_object
```

```{r}
str(character_object)
```
*class in this case is confusing as that is also the type - actually class is character vector in this case*
```{r}
class(character_object)
```

```{r}
attributes(character_object)
```

```{r}
character_object2 <- c("A", "B", "C")
character_object2
```

```{r}
str(character_object2)
```

```{r}
class(character_object2)
```

```{r}
attributes(character_object2)
```

# Exercise

2. Factors 
*I think the class of this is also a vector (special type) - its just factor vector*
```{r, include=TRUE}
factor_object <- factor(c("A", "B", "C"))
factor_object
```

```{r}
str(factor_object)
```

```{r}
class(factor_object)
```

```{r}
attributes(factor_object)
```

```{r}
attr(factor_object, "levels")
```

```{r}
attr(factor_object, "class")
```

# Exercise

3. Data-frames
*data-frame is a special kind of list where it requires all vectors of list to be of same length - so that it can put them in a table*
it fills by column as that is default in R; otherwise you have specify it.
```{r}
dataframe_object <- data.frame(
  numbers = c(1, 2, 3),
  strings = c("A", "B", "C")
  )
dataframe_object
```

```{r}
class(dataframe_object)
```

```{r, include=TRUE}
str(dataframe_object)
```

```{r}
attributes(dataframe_object)
```

# Exercise

4. Matrices
*Matrix is a vector with 2D; it is very different from df*
```{r}
matrix_object <- matrix(
  data = c(1, 2, 3, 4, 5, 6, 7, 8),
  nrow = 2,
  )
matrix_object
```
*it has both classes as matrix is an array with 2D*
```{r}
class(matrix_object)
```

```{r}
str(matrix_object)
```

```{r}
attributes(matrix_object)
```

```{r}
attr(matrix_object, "dim")
```

# Exercise

5. S3 method dispatch
*do not add '.' to your variable names as it has a special meaning in R, thats why use chr, nums and '_'*
```{r}
methods("print") %>% head(14)
```

```{r}
print.factor(factor_object)
```

```{r}
print.factor(character_object)
```

```{r}
print.default(character_object)
```

```{r}
class(NA)
```

